Tulai

### 2024-10-16 paper/paper.qmd
log-transform data before PCA to normalise

---
title: ""
author: ""
date: now
format: 
  html:
    self-contained: true
execute: 
  warning: false
  fig-dpi: 600
title-block-published: "last updated"
---

@@

@@

ggcorrplot(chisq$residuals,
           method = 'circle') + 
  scale_y_discrete(labels = function(y) str_to_upper(y)) +  
  scale_x_discrete(labels = function(x) str_to_lower(x)) +   
  ggplot2::scale_fill_gradient2(
    low = 'red',
    high = 'royalblue',
    mid = 'white',
    midpoint = 0,
    limit = c(-7, 11),
    space = "Lab",
    name = "Residual"
  ) + 
  coord_flip()

@@

ggcorrplot(chisq$residuals,
           method = 'circle') + 
  scale_y_discrete(labels = function(y) str_to_upper(y)) +  # Convert x axis labels to uppercase
  scale_x_discrete(labels = function(x) str_to_lower(x)) +  # Convert x axis labels to uppercase
  ggplot2::scale_fill_gradient2(
    low = 'red',
    high = 'royalblue',
    mid = 'white',
    midpoint = 0,
    limit = c(-7, 11),
    space = "Lab",
    name = "Residual"
  ) + 
  coord_flip()

@@

@@

WABI_Level <- ggplot(filtered_data,  
            aes(x = lithic_density,
                y = retouch_frequency)) +
  geom_point(aes(color = excavation_area,
                 shape = excavation_area), 
             size = 4) +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "blue", 
              linewidth = 0.5) +
  # Add text labels based on 'label' column
  # BM: this is useful for investigating and
  # preparing, but not really informative 
  # for the reader, so we omit for publication
  # geom_text_repel(aes(label = label2)) +
  stat_poly_eq(aes(label = paste(stat(eq.label), 
                                  stat(rr.label), 
                                  stat(p.value.label), 
                                  sep = "~~~")),
               formula = y ~ x, 
               parse = TRUE, 
               size = 5,
               label.x = "right",
               label.y = "bottom") +
  labs(
    x = "Lithic Volumetric Density",
    y = "Retouch Frequency",
    colour = "Excavation area",
    shape = "Excavation area"
  ) +
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  theme(
    legend.position = c(.8,.3),
    axis.text = element_text(size = 7),
    axis.title = element_text(size = 10),
    plot.title = element_text(size = 12)
  )

@@

pca_data <- 
sum_depth_sheet_subset %>% 
  select(lithic_density, 
         core_frequency, 
         blank_frequency, 
         retouch_frequency, 
         tool_diversity)

@@

```{r}
# PCA test
# BM: new method to me! I found https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8858582/

library(PCAtest)

@@

contribution_plots <- 
map(1:4, ~fviz_contrib(res.pca, 
                   choice = "var", 
                   axes = .x, 
                   top = 10))

#combine individual plots into one plot 
library(cowplot)
plot_grid(plotlist = contribution_plots,
                   ncol = 2)

@@

names(pca_data) <- str_replace_all(names(pca_data), "_", " ")

@@

@@

# Which specific pairs of groups differ from each other?
# https://github.com/pmartinezarbizu/pairwiseAdonis
library(pairwiseAdonis)
pairwise.adonis2_results <-
pairwise.adonis2(pca_data[, 1:5] ~ excavation_area,
                data = pca_data_for_post_hoc_testing)

>>>>

---
title: "Tulai"
author:
  - name: First Author
    orcid: 0000-0000-0000-0000
  - name: Second Author
    orcid: 0000-0000-0000-0001
date: now
format: 
  html:
    self-contained: true
    title-block-published: "last updated"
execute: 
  warning: false
  fig-dpi: 600

---

The manuscript with figures generated here is at this URL: https://docs.google.com/document/d/1t7oICQn-sWIFlSJSL92ZVy-iKNMAE7kAJ6MvFc0vKVI/edit?usp=sharing

@@

```{r}
# we need this pkg for the inline R code below this chunk
# devtools::install_github('achetverikov/apastats')

@@

artefact_types_corrplot <- 
ggcorrplot(chisq$residuals,
           method = 'circle') + 
  scale_y_discrete(labels = function(y) str_to_upper(y)) +  
  scale_x_discrete(labels = function(x) str_to_lower(x)) +   
  ggplot2::scale_fill_gradient2(
    low = 'red',
    high = 'royalblue',
    mid = 'white',
    midpoint = 0,
    limit = c(-7, 11),
    space = "Lab",
    name = "Residual"
  ) + 
  coord_flip() +
  guides(fill = "none") +
  theme(axis.text.x  = element_text(angle = 0,
                                    hjust = 0.5))

@@

tool_types_corrplot <- 
ggcorrplot(chisq$residuals,
           method = 'circle') + 
  # Convert x axis labels to uppercase
  scale_y_discrete(labels = function(y) str_to_upper(y)) +  
  # Convert x axis labels to uppercase
  scale_x_discrete(labels = function(x) str_to_lower(x)) +  
  ggplot2::scale_fill_gradient2(
    low = 'red',
    high = 'royalblue',
    mid = 'white',
    midpoint = 0,
    limit = c(-7, 11),
    space = "Lab",
    name = "Residual"
  ) + 
  coord_flip() +
  theme(axis.text.x  = element_text(angle = 0,
                                    hjust = 0.5))

@@

```{r}
# combine corrplots
library(cowplot)

plot_grid(
  artefact_types_corrplot,
  tool_types_corrplot,
  labels = "AUTO"
  )

ggsave(here::here("figures/fig-chi-sq-residuals-both_plots.jpg"),
       h = 6, 
       w = 8)

@@

WABI_Level <- ggplot(filtered_data,  
            aes(x = lithic_density,
                y = retouch_frequency)) +
  geom_point(aes(color = excavation_area,
                 shape = excavation_area), 
             size = 5) +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "blue", 
              linewidth = 0.5) +
  # Add text labels based on 'label' column
  # BM: this is useful for investigating and
  # preparing, but not really informative 
  # for the reader, so we omit for publication
  # geom_text_repel(aes(label = label2)) +
  stat_poly_eq(aes(label = paste(stat(eq.label), 
                                  stat(rr.label), 
                                  stat(p.value.label), 
                                  sep = "~~~")),
               formula = y ~ x, 
               parse = TRUE, 
               size = 5,
               label.x = "right",
               label.y = "top") +
  labs(
    x = "Lithic Volumetric Density",
    y = "Retouch Frequency",
    colour = "Excavation area",
    shape = "Excavation area"
  ) +
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  theme(
    legend.position = c(.8,.3),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 14)
  )

@@

pca_data <- 
sum_depth_sheet_subset %>% 
   mutate(lithic_density = log(lithic_density + 0.001),  # Adding 1 to avoid log(0)
         core_frequency = log(core_frequency + 0.001),
         blank_frequency = log(blank_frequency + 0.001),
         retouch_frequency = log(retouch_frequency + 0.001),
         tool_diversity = log(tool_diversity + 0.001)) %>%

  select(`lithic\ndensity` = lithic_density, 
         `core\nfrequency` = core_frequency, 
         `blank\nfrequency` = blank_frequency, 
         `retouch\nfrequency` = retouch_frequency, 
         `tool\ndiversity` = tool_diversity)

@@

```{r}
# PCA test
# BM: new method to me! I found https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8858582/
# devtools::install_github("arleyc/PCAtest")
library(PCAtest)

@@

contribution_plots <- 
map(1:4, ~fviz_contrib(res.pca, 
                   choice = "var", 
                   axes = .x, 
                   top = 10) +
  theme(axis.text =  element_text(size = 5),
        axis.text.x = element_text(size = 8,
                                   angle = 0,
                                   hjust = 0.5, 
                                   vjust = 1),
        axis.title = element_text(size = 6),
        plot.title = element_text(size = 6)))

#combine individual plots into one plot 
library(cowplot)
contribution_plots_panel <- 
plot_grid(plotlist = contribution_plots,
                   ncol = 1)

@@

names(pca_data) <- str_replace_all(names(pca_data), "\n", " ")

@@

```{r}
# combine PCA plots
plot_grid(biplot_modified,
          contribution_plots_panel,
          ncol = 1,
       labels = "AUTO")

ggsave(
       filename = here::here("figures/fig-pca-biplot-and-contribution.jpg"),
       h = 10, 
       w = 6)

@@

# Which specific pairs of groups differ from each other?
# https://github.com/pmartinezarbizu/pairwiseAdonis
# install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")
library(pairwiseAdonis)
pairwise.adonis2_results <-
pairwise.adonis2(pca_data[, 1:5] ~ excavation_area,
                data = pca_data_for_post_hoc_testing)
